---
title: "My first plot"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message = FALSE)
```

## Get the data

This is a markdown chunk. Chunks always start and end with 3 backticks. You can "play" the chunk right in RStudio or "knit" the whole document together. 

```{r}
# these are comments
# data source: NY Times, https://tinyurl.com/t9j9fdw
# https://github.com/nytimes/covid-19-data

# create an object called nyt_st_url that contains the url
  nyt_st_url <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv"
```

Hit the play button (triangle) and look at the Environment tab. You should see a new object called `nyt_st_url`. Paste `nyt_st_url` (without the backticks) into the console and press return.

## Get the data

Next we'll use a function called `read.csv()` to read in the `csv` data from the website. Hit play to run the chunk or put your cursor on the line and hit "Run".

```{r}
  nyt_st <- read.csv(nyt_st_url, stringsAsFactors = FALSE)
```

You should see the `nyt_st` object in the Environment. Open it by expanding the blue arrow and then clicking on the object name. You'll see a Viewer open as a new tab. Note all of the things you can do in the viewer. There are `r nrow(nyt_st)` observations and `r length(nyt_st)` variables.

[Wondering why I'm using all of these backticks and placeholders?? You'll see soon enough!]

## Load packages

R comes with lots of 'base' packages. Users like you have contributed more than 15,000 additional packages to a repository called CRAN. You can find still more packages on sites like GitHub.

```{r}
# install packages only once on your machine 
# if running on your own computer:
#   copy/paste lines 47-49 in Console
#   don't include the # which tells R to ignore rest of line

# install.packages(c("tidyverse", "devtools", "ggrepel", "ggthemes",
#                    "knitr", "kableExtra"))
# devtools::install_github("wilkelab/ggtext")

# load necessary packages every time you open your R session
  library(tidyverse)
  library(ggtext)
  library(ggrepel)
  library(ggthemes)
  library(knitr)
  library(kableExtra)
```

## The pipe %>% operator 

Read from left to right, top to bottom. Speak `%>%` as "then". `nyt_st %>% group_by(state)` is "start with data then do_something() to it". Hit play to count observations there are by `state`.

```{r}
  nyt_st %>%
    group_by(state) %>%
    count() 
```

## Make the output a bit nicer

We can use `knitr::kable()` (which means the `kable()` function in the `{knitr}` pacakge) and `kableExtra::kable_styling()` to format the tibble output.

```{r}
  nyt_st %>%
    group_by(state) %>%
    count() %>%
    kable() %>% 
    kable_styling()
```


## An example of data wranging

Even "clean" data files are rarely ready for plotting out of the box. We'll unpack the specifics of what this code does as we move through the workshop series. For now, run the chunk.

```{r, results='hide'}
  nyt_st %>%
    group_by(state) %>%
    mutate(dailyDeaths = c(deaths[1], diff(deaths))) %>%
    filter(dailyDeaths >= 3) %>%
    mutate(days = 1:n()) %>%
    mutate(label = if_else(days == max(days), 
                           state, 
                           NA_character_)) %>%
    ungroup() %>%
    filter(state == "Florida" | 
           state == "New York" |
           state == "Washington")
```

No output is printed because we set the chunk options to hide.

## Assign the output to an object

This time we're going to run the same chunk but we'll assign the output to a new object called `df_deaths`. You can name objects as you wish. `df_deaths` is snake case. I also like `namingWithCamelCase`.

```{r}
  df_deaths <- # this is new!
  nyt_st %>%
    group_by(state) %>%
    mutate(dailyDeaths = c(deaths[1], diff(deaths))) %>%
    filter(dailyDeaths >= 3) %>%
    mutate(days = 1:n()) %>%
    mutate(label = if_else(days == max(days), 
                           state, 
                           NA_character_)) %>%
    ungroup() %>%
    filter(state == "Florida" | 
           state == "New York" |
           state == "Washington")
```

You should see `df_deaths` in your Environment. You can re-use it for later analyses.

## `ggplot()`

We're going to dig into `ggplot()` in our next session. Let's hit a few of the highlights. What happens when you just call the function on the data?

```{r, out.width='70%'}
# out.width='70%' makes it possible to fit the plot on the slide
  ggplot(df_deaths)
```

## Aesthetic mappings

With `aes()` we tell ggplot how our variables map onto plot features. For instance, `days` will go on the x-axis.

```{r, out.width='70%'}
  ggplot(df_deaths, aes(x=days, y=deaths, group=state, color=state))
```

## Add a geom

Next we'll tell `ggplot()` what type of geometric object to add. Let's add lines.

```{r, out.width='70%'}
  ggplot(df_deaths, aes(x=days, y=deaths, group=state, color=state)) +
    geom_line() 
```

## We have more to add, but first...

We don't have to keep repeating code over and over. We can assign what we have so far to an object, let's call it `p`, and then add to `p`. 

```{r}
  p <- ggplot(df_deaths, aes(x=days, y=deaths, group=state, 
                             color=state)) +
        geom_line()

  p <- p + 
    scale_y_log10(breaks = c(1, 2, 5, 10, 20, 50, 100, 
                             200, 500, 1000, 2000, 5000, 10000)) +
    scale_x_continuous(breaks = c(1, 5, 10, 15, 20, 25, 30),
                       limits = c(1, 30)) +
    geom_text_repel(aes(label = label),
                     nudge_x = 1,
                     na.rm = TRUE) 
```

## Make it fancier

You can lose hours and days making plots perfect. Let's use the FiveThirtyEight theme and wrap up.

```{r}
  p <- p + 
    theme_fivethirtyeight() +
    theme(plot.title.position = "plot",
          plot.caption = element_markdown(),
          legend.position = "none",
          axis.title.x = element_text()) + 
    labs(title = "Put your main headline here",
         subtitle = "Daily deaths with coronavirus, by number of days 
                     since 3 daily deaths first recorded",
         x = "Number of days since 3 daily deaths first recorded", 
         caption = "Data from *The New York Times*, based on reports 
                    from state and local health agencies.")
```

## Final product

```{r, echo=FALSE}
# echo=FALSE in the chunk options hides the code
# print the plot
  p
```

