---
title: "Week 5"
description: |
  Let's make some interactive and animated plots!
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Interactive plot with `{plotly}`

## Back to Week 1 for the static version

```{r static, echo=FALSE}
# your project is missing a few packages, so run all commented
  library(tidyverse)
  library(plotly)
  library(ggthemes)
  library(ggrepel)
  library(ggtext)
  library(gganimate) # install.packages("gifski")
  library(viridis)
  library(ggflags) # devtools::install_github("ellisp/ggflags")
  library(countrycode) # install.packages("countrycode")

  nyt_st <- read.csv("https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-states.csv", stringsAsFactors = FALSE)
  
  p <- 
  nyt_st %>%
    group_by(state) %>%
    mutate(dailyDeaths = c(deaths[1], diff(deaths))) %>%
    filter(dailyDeaths >= 3) %>%
    mutate(days = 1:n()) %>%
    mutate(label = if_else(days == max(days), 
                           state, 
                           NA_character_)) %>%
    ungroup() %>%
    filter(state == "Florida" | 
           state == "New York" |
           state == "Washington") %>%
    ggplot(., aes(x=days, y=deaths, group=state, 
                             color=state)) +
      geom_line() +
      scale_y_log10(breaks = c(1, 2, 5, 10, 20, 50, 
                               100, 200, 500, 1000, 
                               2000, 5000, 10000)) +
      scale_x_continuous(breaks = c(1, 5, 10, 15, 20, 25, 30),
                         limits = c(1, 30)) +
      geom_text_repel(aes(label = label),
                      nudge_x = 1,
                      na.rm = TRUE) +
      theme_fivethirtyeight() +
      theme(plot.title.position = "plot",
            plot.caption = element_markdown(),
            legend.position = "none",
            axis.title.x = element_text()) + 
      labs(title = "Put your main headline here",
           subtitle = "Daily deaths with coronavirus, by number of days since 3 daily deaths first recorded",
           x = "Number of days since 3 daily deaths first recorded", 
           caption = "Data from *The New York Times*, based on reports from state and local health agencies.")
      
  p
```

## Make interactive with `plotly::ggplotly()`

```{r ggplotly}
  ggplotly(p)
```

That's it.

# Animated plot with `{gganimate}`

## Our inspiration

Our inspiration comes from Bobbie Johnson in [*MIT Technology Review*](https://www.technologyreview.com/2020/03/27/950263/the-covid-19-pandemic-in-two-animated-charts/).

<!-- I just copied/pasted Twitter embed code -->

<blockquote class="twitter-tweet"><p lang="en" dir="ltr">What are your favorite animated <a href="https://twitter.com/hashtag/COVID19?src=hash&amp;ref_src=twsrc%5Etfw">#COVID19</a> charts? Tomorrow is Week 5 of &quot;I Eat <a href="https://twitter.com/hashtag/DataScience?src=hash&amp;ref_src=twsrc%5Etfw">#DataScience</a> for Breakfast: Pandemic Ed.&quot; and I want to cover animated charts. Looking for a good one to reproduce in <a href="https://twitter.com/hashtag/RStats?src=hash&amp;ref_src=twsrc%5Etfw">#RStats</a>.<a href="https://t.co/1h50KIj0HR">https://t.co/1h50KIj0HR</a><br><br>Contender from <a href="https://twitter.com/bobbie?ref_src=twsrc%5Etfw">@bobbie</a> (<a href="https://t.co/tFOhM4hPtW">https://t.co/tFOhM4hPtW</a>) <a href="https://t.co/BDcfZ4HVtu">pic.twitter.com/BDcfZ4HVtu</a></p>&mdash; Eric Green (@ericpgreen) <a href="https://twitter.com/ericpgreen/status/1255848835865350144?ref_src=twsrc%5Etfw">April 30, 2020</a></blockquote> <script async src="https://platform.twitter.com/widgets.js" charset="utf-8"></script>

## Setup

Just as in Week 3, the data for today's plot come from the [JHU Center for Systems Science and Engineering data repository](https://github.com/CSSEGISandData/COVID-19).

```{r load}
# get the data
# source: https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series
  
  cases_wide <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv", stringsAsFactors = FALSE)
  
# show a small slice of the data
  cases_wide %>%
    slice(1:10) %>%
    select(Province.State, Country.Region, Lat, Long, X1.22.20) %>%
    knitr::kable() %>%
    kableExtra::kable_styling()
```

## Wrangle the data from wide to long

The data come to us in a wide format with column after column of daily culumative case counts. We need to pivot this table to a long version, but you might also recall that we need to aggregate some sub-national data to the national level. We did this all before in Week 3.

```{r tolong}
  cases <- 
  cases_wide %>%
  # sum subnational data to get national totals
    select(-Lat, -Long) %>%                 # drop these columns
    group_by(Country.Region) %>%            # group by country
    mutate_at(vars(starts_with("X")),       # sum each daily column
              funs(sum)) %>%                #  by county
    ungroup() %>%
    distinct(Country.Region,                # countries with subnational
             .keep_all = TRUE) %>%          #  have repetitive data now 
    select(-Province.State) %>%             #  so drop duplicates
  # pivot wide to long
    pivot_longer(cols = starts_with("X"),   # take the daily cases
                 names_to = "date",         #  put col headings in date
                 names_prefix = "X",        #  pattern is X variables
                 values_to = "cases") %>%   #  put case values in case
    mutate(date = lubridate::mdy(date)) %>% # create proper date var
    rename("country"="Country.Region") %>%  # give nicer var name
    filter(country!="Diamond Princess")     # drop diamond princess
    
# show a small slice of the data
  cases %>%
    slice(1:10) %>%
    knitr::kable() %>%
    kableExtra::kable_styling()
```


## Limit to 10 country-observations per day

The original plot shows the daily leaderboard in the race to the most cases. Our dataset has every country, every day. Let's limit it to the top 10 countries every day by filtering out some rows. We'll follow [this nice approach by Joe Tran](https://towardsdatascience.com/https-towardsdatascience-com-everything-you-need-to-know-about-animated-bar-charts-be033f398619).

```{r top10}
  leaderboard <- 
  cases %>%
    group_by(date) %>%
    # mutate(rank = min_rank(-cases) * 1,
    #        relative_to_1 = cases / cases[rank == 1],
    #        label = paste0(" ", cases)) %>%
    mutate(rank = rank(-cases),
           relative_to_1 = cases/cases[rank == 1],
           label = paste0(" ", cases)) %>%
  # lots of countries tied for zero at the start
  # so order by rank and then alphabetical by country
    arrange(rank, country) %>%
  # give new ranks to break ties with alphabetical country names
    mutate(rank = 1:n()) %>%
  # limit to top 10 per day
    filter(rank <= 10) %>%
    ungroup() %>%
  # give a day label
    mutate(month = lubridate::month(date, label=TRUE, abbr=TRUE),
           day = lubridate::day(date),
           monthDay = paste(month, day, sep=" ")) %>%
  # convert country name to code
    mutate(cc = countrycode(country, origin="country.name", 
                            destination = "genc2c"),
           cc = tolower(cc))

# get vector of unique monthDay
  monthDayLevels <- 
  leaderboard %>% 
    distinct(date, .keep_all = TRUE) %>% 
    pull(monthDay)
```

## Create the static plot

Before we can animate the plot we need to create the static version.

```{r ggplot}
  static <- 
  leaderboard %>%
  # convert to factor
    mutate(monthDay = factor(monthDay, 
                             levels = monthDayLevels,
                             labels = monthDayLevels)) %>%
  # plot
    ggplot(., aes(x = rank, group = country, 
                  fill = country, color = country,
                  country = cc)) +
      geom_tile(aes(y = cases/2, height = cases,
                       width = 0.9), alpha = 0.8, color = NA) +
      #geom_flag(aes(y = cases+50), size = 20) +
      geom_text(aes(y = 0, label = paste(country, "   ")),
                    vjust = 0.2, hjust = 1, size = 10) +
      geom_text(aes(y=cases+50, label = label, hjust=0),
                    size = 10) +
      coord_flip(clip = "off", expand = TRUE) +
      scale_x_reverse() +
  
    # making it look nice
      scale_fill_viridis_d(option = "plasma") +
      scale_color_viridis_d(option = "plasma") +
      theme_minimal() +
      theme(axis.line=element_blank(),
            axis.text.x=element_blank(),
            axis.text.y=element_blank(),
            axis.ticks=element_blank(),
            axis.title.x=element_blank(),
            axis.title.y=element_blank(),
            legend.position="none",
            panel.background=element_blank(),
            panel.border=element_blank(),
            panel.grid.major=element_blank(),
            panel.grid.minor=element_blank(),
            panel.grid.major.x = element_line(size=.4, 
                                              color="grey" ),
            panel.grid.minor.x = element_line(size=.1, 
                                              color="grey" ),
            plot.title.position = "plot",
            plot.title=element_text(size=20, 
                                    face="bold", 
                                    colour="#313632"),
            plot.subtitle=element_text(size=50, 
                                       color="#a3a5a8"),
            plot.caption =element_text(size=15, 
                                       color="#313632"),
            plot.background=element_blank(),
            plot.margin = margin(1, 9, 1, 9, "cm")) +
      labs(title = 'Cumulative number confirmed COVID-19 cases by date and country',  
           subtitle = '{closest_state}',
           caption  = "Data: Johns Hopkins CCSE, https://tinyurl.com/qooo4fp")

static
```

## Define transition states and easing

In this next chunk we will tell `{gganimate}` which variable segments the plot into different frame and define the nature of the transitions and easing.

```{r}
  animate <- 
  static +
    #transition_time(date) +
    transition_states(monthDay,
                      transition_length = 4,
                      state_length = 1) +
    ease_aes('cubic-in-out') +
    view_follow(fixed_x = TRUE) 
```

## Create a gif

In the final step we'll use the `animate()` function to render the plot. This will take a while.

```{r, message=FALSE, warning=FALSE, eval=TRUE}
  animate(animate, fps = 10, duration = 50, width = 1000, height = 600,
          renderer=av_renderer("img/gganim.mp4"))
```









